MLH1 Results Outline

### 2020-02-21 doc/Results.rmd
clean up and org for Q2 SC length

@@

@@

```{r juvimice.list.copy, echo=FALSE}
juvi_mice <- c(
  "30dec14_WSB_m2","30dec14_WSB_m4","30dec14_WSB_m1","30dec14_WSB_m3","30dec14_WSB_m5",
  "30sep14_G_m1","30sep14_G_m2","30sep14_G_m3","30sep14_G_m4","30sep14_G_m5",
  "12Feb17_LEW_m1","12Feb17_LEW_m2", "13sep17_LEW_m2", "13sep17_LEW_m1","13sep17_LEW_m3",

@@

                        #mean.first.1CO
                        norm.first.foci.pos =  mean(PER_Foci_1[which(hand.foci.count == 1)])

@@

First we examine the pattern of SC length variation across chromosome classes.

@@

cell.avlong.biv <- ggplot(cell.av.long.biv.table, aes(y=mean.sc, x=strain, color=strain))+geom_jitter()+facet_wrap(~subsp, scales = "free")+scale_color_manual(values=colors_of_strains)+ylim(c(75,140))+geom_boxplot(alpha=.2) + ggtitle("Cell av. \nMale Long Biv SC Lengths")
mouse.shoo.long.biv
```

```{r Q2.SCshort_plot, echo=FALSE}

@@

mouse.av_totsc_male

```{r Q2_SC_ChrmClass_long.biv, echo=FALSE}
#there might not be enough data 
#
ii <- as.data.frame( table(male.long.biv$hand.foci.count, male.long.biv$strain) )
```

total SC for males (which ar)
X additional cells were added to the male specific total SC data set described above

<!-- tests below;  t.test(high low) OR logistic regression, mixed modlels, glm 
remember the pairwise t.test
Pwise.ttest_Dom.male <- pairwise.t.test(male.Dom.mouse.av$mean_SC, male.Dom.mouse.av$strain)
G-WSB is sig, WSB-LEW are slightly sig

```{r Q2.totSC.ttest, echo=FALSE}

#totalSC high and low rec group

#
Q2tot.sc_t.test_High.Low <- t.test((totSC.mouse.av.table %>% filter(sex == "male") %>% filter(Rec.group == "0") )$mean.skel,

                               (totSC.mouse.av.table %>% filter(sex == "male") %>% filter(Rec.group == "1") )$mean.skel)
#p-value = 0.007

@@

**we ran t.tests on the SC based metrics to test is the high and low groups are significanly different**

**mouse means of total SC area for Higher rec strains are significantly higher compared to low rec strains (ttest; p = 
`r Q2tot.sc_t.test_High.Low$p.value`).**

@@

this might indicate -there's error in the total SC algorithm

- mouse effects

- low power / sampling bias in reduced data set

- total SC is capturing differences in the XY signal which is excluded from the bivalent data set

glms

@@

For M2 all

@@

**general pattern: SC lengths correlate with number of foci on single bivalents (the chromosome classes). Diviations within strains are due to low sample numbers.**

We purpose -- we think that running the models for comparing SC means by chromosome class and divided by rec group. (since the chromosome proportions are similar -- hopefully/we assume the the sets of chromosomes within each set are more similar )

@@

The skeletonized measures show that the general pattern holds for total SC and the high rec strains. However there are a few exceptions -- However in the molf strain, there might be strain specific evolution--the mol subsp all have more SC skel

@@

The general pattern is that the high rec strain have a greater mean SC length for all pooled bivalents (`r length(Curated_BivData$Obj.ID[Curated_BivData$sex == "male"])` males bivalent observations).

The predicted pattern for SC length becomes more nuanced when the data are split up by Chromosome class. Mainly that 1CO are shorter in the high rec strains then low rec strains. This is likely due to the fact that more physically longer chromosomes have 1CO in the low rec strains, which pushes the mean SC length up. Where as in the High rec strains, physically longer chromosomes are more likely to be in the 2CO group.  This supports a general pattern of tighter clustering of SC lengths across chromosome classes in the high rec group. They have a lower probability of 2COs below a certain SC length threshold.

@@

```{r male.poly.test.MM, echo=FALSE, eval=FALSE}
#I don't think the Mixed models fit well for the high vs low polymorphism tests.

#t.test comparisons of PWD_SC vs (KAZ, SKIVE, )
#MSM_SC vs MOLF_SC

#2. logistic regression for sc length and chrm class?
#if the prediction / significance

#isolate data
#mouse.avs_4MM_male <- mouse.avs_4MM[mouse.avs_4MM$sex == "male",]
#mouse.avs_4MM_M.male <- mouse.avs_4MM_male[mouse.avs_4MM_male$subsp == "Musc",]

mouse.avs_4MM_M.male$mouse <- factor(mouse.avs_4MM_M.male$mouse, ordered = FALSE)

#build and test MM
lme_MM_SC_length_male.poly <- lme(mean_SC ~ mouse + 1|strain, data= mouse.avs_4MM_M.male)#this syntax not working
as.data.frame( round(  summary(lme_MM_SC.length)$tTable, 6 ) )

lmer_MM_SC_length_male.poly <- lmer(mean_SC ~ (1|strain), data=mouse.avs_4MM_M.male, REML = TRUE)

anova(lmer_MM_SC_length_male.poly)

rand_SC.Length_MM.male.poly <- exactRLRT(lmer_MM_SC_length_male.poly)#this won't work either...hmm

@@

Each point is a mouse average SC. When all bivalents are pooled, the logistic regression SC lengths predict the groups. When the data is divided by CO number, 1CO's have a higher SC length in the low group - in the low group more chrms, including the longer ones have 1CO - whereas in the high group, only shorter chrms (SC) are in the 1CO class.
For the 2CO class of bivalents - all the SC are longer, the high group averages were longer than the low.

 and the mean 2COs SC are higher in the high strains
(this must have something to do with the physical size effect)

If I want to try to use mouse nested in strain -- I should use cell level metrics (but those have there own flaws)

Ideas for the above 3 tests,

1. no difference in DOM
- no sig logistic regression test
- glm, strain effect?

>>>>

# transition Q1 to Q2

we tested the sex differences -- now we focus on the largest aspect of variation, the high rec males vs the low rec males.

@@

motivation:
the positive correlation, Our prediction for this metric is that the high rec strain will have longer sc lengths (axis lengths)

<pattern for the single bivalent, pos correlation>
The predicted pattern for SC length becomes more nuanced when the data are split up by Chromosome class. Mainly that 1CO are shorter in the high rec strains then low rec strains. This is likely due to the fact that more physically longer chromosomes have 1CO in the low rec strains, which pushes the mean SC length up. Where as in the High rec strains, physically longer chromosomes are more likely to be in the 2CO group.  This supports a general pattern of tighter clustering of SC lengths across chromosome classes in the high rec group. They have a lower probability of 2COs below a certain SC length threshold.

<NEW STRAINS INTRODUCED NOT IN PREV DATA>

this section / these tests follow up with those made in Q1 which found that the male musc and mol total-SC were signifcantly higher compared to the male Dom total SC measures.
These test include additional measues from strains which only had male observations.

total SC for males (which are), X additional cells were added to the male specific total SC data set described above

@@

```{r Q2.SC.juvimice.list.copy, echo=FALSE}
juvi_mice <- c(
  "30dec14_WSB_m2","30dec14_WSB_m4","30dec14_WSB_m1","30dec14_WSB_m3","30dec14_WSB_m5",
  "30sep14_G_m1","30sep14_G_m2","30sep14_G_m3","30sep14_G_m4","30sep14_G_m5",
  "12Feb17_LEW_m1","12Feb17_LEW_m2", "13sep17_LEW_m2", "13sep17_LEW_m1","13sep17_LEW_m3",

@@

                        #mean.first.1CO
                        norm.first.foci.pos =  mean(PER_Foci_1[which(hand.foci.count == 1)]
                         )

@@

<!-- plots -->

@@

cell.avlong.biv <- ggplot(cell.av.long.biv.table, aes(y=mean.sc, x=strain, color=strain))+geom_jitter()+facet_wrap(~subsp, scales = "free")+scale_color_manual(values=colors_of_strains)+ylim(c(75,140))+geom_boxplot(alpha=.2) + ggtitle("Cell av. \nMale Long Biv SC Lengths")
```

@@

mouse.av_totsc_male
```

<!-- t tests-->
<!-- tests below;  t.test(high low) OR logistic regression, mixed modlels, glm 
remember the pairwise t.test
Pwise.ttest_Dom.male <- pairwise.t.test(male.Dom.mouse.av$mean_SC, male.Dom.mouse.av$strain)
G-WSB is sig, WSB-LEW are slightly sig
-->

```{r Q2.totSC.ttest, echo=FALSE}
#totalSC high and low rec group

Q2tot.sc_t.test_High.Low <- t.test((totSC.mouse.av.table %>% filter(sex == "male") %>% filter(Rec.group == "0") )$mean.skel,
                (totSC.mouse.av.table %>% filter(sex == "male") %>% filter(Rec.group == "1") )$mean.skel)
#p-value = 0.007

#mol
Q2tot.sc_t.test_High.Low_Mol <- t.test((totSC.mouse.av.table %>% filter(sex == "male") %>% filter(Rec.group == "0") %>% filter(subsp == "Mol") )$mean.skel,
         (totSC.mouse.av.table %>% filter(sex == "male") %>% filter(Rec.group == "1") %>% filter(subsp == "Mol") )$mean.skel)

#musc
Q2tot.sc_t.test_High.Low_Musc <- t.test((totSC.mouse.av.table %>% filter(sex == "male") %>% filter(Rec.group == "0") %>% filter(subsp == "Musc"))$mean.skel,
                (totSC.mouse.av.table %>% filter(sex == "male") %>% filter(subsp == "Musc") %>% filter(Rec.group == "1") )$mean.skel)

@@

total SC t test

 p = `r Q2tot.sc_t.test_High.Low$p.value`
**mouse means of total SC area for Higher rec strains are significantly higher compared to low rec strains (ttest; p = 
`r Q2tot.sc_t.test_High.Low$p.value`).**

musc and mol ttests

p = `r`

p= `r`

Additionally we ran similar tests using the short and long bivalent data sets which help to alimeorate potential chromosome size effects.

@@

<this might indicate -there's error in the total SC algorithm
- mouse effect
- low power / sampling bias in reduced data set
- total SC is capturing differences in the XY signal which is excluded from the bivalent data set>

<!-- models -->

@@

@@

<we ran models to start examine strain specific effects ... modeling the fixed effects of subspeices and strain.>

Thse are our prediction for the glm models -- based on the variation of the gwRR's

1. no difference in DOM strains (for SC lengths (CONFIRM IF THERE ARE DIFFERENECES FOR GWRR))

2. PWD > SKIVE > the rest

3. Msm > mol

We purpose we think that running the models for comparing SC means by chromosome class and divided by rec group. (since the chromosome proportions are similar -- hopefully/we assume the the sets of chromosomes within each set are more similar )

@@

The skeletonized measures show that the general pattern holds for total SC and the high rec strains. However there are a few exceptions -However in the molf strain, there might be strain specific evolution--the mol subsp all have more SC skel

@@

@@

@@

<Each point is a mouse average SC. When all bivalents are pooled, the logistic regression SC lengths predict the groups. When the data is divided by CO number, 1CO's have a higher SC length in the low group - in the low group more chrms, including the longer ones have 1CO - whereas in the high group, only shorter chrms (SC) are in the 1CO class.
For the 2CO class of bivalents - all the SC are longer, the high group averages were longer than the low.>
